---
title: "Exploratory Data Analysis"
output: html_document
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=4)
library(plyr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(grid)
library(choroplethr)
library(choroplethrMaps)
```

## Project Data

Let's start with exploratory data analysis of the project-specific data. This includes information on school location, school type, teacher attributes, project categories, project pricing and impact, project donations, and project status.

#### Importing the file
```{r File Import, cache=TRUE}
project_orig <- read.csv(gzfile("opendata_projects000.gz"), sep = ",", header = FALSE, col.names = c('_projectid', '_teacher_acctid', '_schoolid', 'school_ncesid', 'school_latitude', 'school_longitude', 'school_city', 'school_state', 'school_zip', 'school_metro', 'school_district', 'school_county', 'school_charter', 'school_magnet', 'school_year_round', 'school_nlns', 'school_kipp', 'school_charter_ready_promise', 'teacher_prefix', 'teacher_teach_for_america', 'teacher_ny_teaching_fellow', 'primary_focus_subject', 'primary_focus_area' ,'secondary_focus_subject', 'secondary_focus_area', 'resource_type', 'poverty_level', 'grade_level', 'vendor_shipping_charges', 'sales_tax', 'payment_processing_charges', 'fulfillment_labor_materials', 'total_price_excluding_optional_support', 'total_price_including_optional_support', 'students_reached', 'total_donations', 'num_donors', 'eligible_double_your_impact_match', 'eligible_almost_home_match', 'funding_status', 'date_posted', 'date_completed', 'date_thank_you_packet_mailed', 'date_expiration'))
```

#### Cleaning the data

```{r Clean Data}
project <- tbl_df(project_orig)
project$X_projectid <- as.character(project$X_projectid)
project$X_teacher_acctid <- as.character(project$X_teacher_acctid)
project$X_schoolid <- as.character(project$X_schoolid)
project$school_ncesid <- as.character(project$school_ncesid)

project$school_zip <- as.factor(project$school_zip)

project$school_charter <- as.logical(toupper(as.character(project$school_charter)))
project$school_magnet <- as.logical(toupper(as.character(project$school_magnet)))
project$school_year_round <- as.logical(toupper(as.character(project$school_year_round)))
project$school_nlns <- as.logical(toupper(as.character(project$school_nlns)))
project$school_kipp <- as.logical(toupper(as.character(project$school_kipp)))
project$school_charter_ready_promise <- as.logical(toupper(as.character(project$school_charter_ready_promise)))
project$teacher_teach_for_america <- as.logical(toupper(as.character(project$teacher_teach_for_america)))
project$teacher_ny_teaching_fellow <- as.logical(toupper(as.character(project$teacher_ny_teaching_fellow)))
project$eligible_double_your_impact_match <- as.logical(toupper(as.character(project$eligible_double_your_impact_match)))
project$eligible_almost_home_match <- as.logical(toupper(as.character(project$eligible_almost_home_match)))

project$date_posted <- as.Date(project$date_posted)
project$date_completed <- as.Date(project$date_completed)
project$date_thank_you_packet_mailed <- as.Date(project$date_thank_you_packet_mailed)
project$date_expiration <- as.Date(project$date_expiration)

project$grade_level <- factor(project$grade_level, levels = c("", "Grades PreK-2", "Grades 3-5", "Grades 6-8", "Grades 9-12"))
project$poverty_level <- factor(project$poverty_level, levels = c("", "low poverty", "moderate poverty", "high poverty", "highest poverty"))

```


#### Funding status
Since the goal of this capstone project is to predict whether or not a project will be successfully funded, let's first take a look at the different options for funding status.

``` {r}
ggplot(project, aes(x = funding_status)) + geom_bar() + coord_flip() + theme_few()
```

We will only focus on the projects that are either completed (successfully funded) or expired (did not reach funding goal), and remove the live and reallocated projects.

``` {r}
project <- project %>% filter(funding_status == "completed" | funding_status == "expired")
```

Next, let's look at some of the properties of each project (type of school, type of project, school location, etc.).

#### Type of School

There are several properties which indicate the type of school, including:

* school_charter
* school_magnet
* school_year_round
* school_nlns (New Leaders for New Schools program)
* school_kipp
* school_charter_ready_promise

Let's look at each of these properties and how they relate to funding status.
``` {r Type of School}
Palette1 <- c("#7AC36A", "#F15A60")
draw_graphs <- function(df,prop1,prop2,name1,name2) {
  plot1 <- ggplot(df, aes(x = prop1)) + geom_bar() + coord_flip() + theme_few() + scale_y_continuous(name = "Total Count") + scale_x_discrete(name = name1)

  plot2 <- ggplot(df, aes(x = prop1, fill = prop2)) + geom_bar(position = "fill") + coord_flip() + theme_few() + theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) + scale_fill_manual(name = name2, values=Palette1) + scale_y_continuous(name = "Relative Count")

  grid.draw(cbind(ggplotGrob(plot1), ggplotGrob(plot2), size = "last"))
}

draw_graphs(project, project$school_charter, project$funding_status, "Charter School", "Funding Status")
```

``` {r}
draw_graphs(project, project$school_magnet, project$funding_status, "Magnet School", "Funding Status")
```

``` {r}
draw_graphs(project, project$school_nlns, project$funding_status, "NLNS School", "Funding Status")
```

``` {r}
draw_graphs(project, project$school_kipp, project$funding_status, "KIPP School", "Funding Status")
```

``` {r}
draw_graphs(project, project$school_charter_ready_promise, project$funding_status, "Charter Ready Promise School", "Funding Status")
```

#### Teacher Attributes

The attributes of the teacher include:

* teacher_prefix
* teacher_teach_for_america
* teacher_ny_teaching_fellow

``` {r Teacher Attributes}
draw_graphs(project, project$teacher_prefix, project$funding_status, "Teacher Prefix", "Funding Status")
```

``` {r}
draw_graphs(project, project$teacher_teach_for_america, project$funding_status, "Teach for America", "Funding Status")
```

``` {r}
draw_graphs(project, project$teacher_ny_teaching_fellow, project$funding_status, "NY Teaching Fellow", "Funding Status")
```


#### Project Categories

The project categories include:

* primary_focus_subject
* primary_focus_area
* secondary_focus_subject
* secondary_focus_area
* resource_type
* poverty_level
* grade_level

``` {r Project Categories}
draw_graphs(project, project$primary_focus_area, project$funding_status, "Primary Focus Area", "Funding Status")
```

``` {r}
draw_graphs(project, project$primary_focus_subject, project$funding_status, "Primary Focus Subject", "Funding Status")
```

``` {r}
draw_graphs(project, project$secondary_focus_area, project$funding_status, "Secondary Focus Area", "Funding Status")
```

``` {r}
draw_graphs(project, project$secondary_focus_subject, project$funding_status, "Secondary Focus Subject", "Funding Status")
```

``` {r}
draw_graphs(project, project$resource_type, project$funding_status, "Resource Type", "Funding Status")
```

``` {r}
draw_graphs(project, project$poverty_level, project$funding_status, "Poverty Level", "Funding Status")
```

``` {r}
draw_graphs(project, project$grade_level, project$funding_status, "Grade Level", "Funding Status")
```


#### Project Pricing and Impact

These fields are available under Project Details on a project page, breaking down the cost of a project into its parts.

* vendor_shipping_charges
* sales_tax
* payment_processing_charges
* fulfillment_labor_materials
* total_price_excluding_opt_donation
* total_price_including_opt_donation
* students_reached

Let's look at some of these properties in more detail:

``` {r}
summary(project$total_price_excluding_optional_support)
summary(project$total_price_including_optional_support)
summary(project$students_reached)

draw_histograms <- function(df,prop1,prop2,name1,name2) {
  plot1 <- ggplot(df, aes(x = prop1)) + geom_histogram() + coord_flip() + theme_few() + scale_y_continuous(name = "Total Count") + scale_x_continuous(name = name1)

  plot2 <- ggplot(df, aes(x = prop1, fill = prop2)) + geom_histogram(position = "fill") + coord_flip() + theme_few() + theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) + scale_fill_manual(name = name2, values=Palette1) + scale_y_continuous(name = "Relative Count")

  grid.draw(cbind(ggplotGrob(plot1), ggplotGrob(plot2), size = "last"))
}

draw_histograms(project, log10(project$total_price_excluding_optional_support), project$funding_status, "Log(Total Price Excluding Optional Support)", "Funding Status")
```

``` {r}
draw_histograms(project, log10(project$total_price_including_optional_support), project$funding_status, "Log(Total Price Including Optional Support)", "Funding Status")
```

``` {r}
draw_histograms(project, log10(project$students_reached), project$funding_status, "Log(Students Reached)", "Funding Status")

```

To Do: Take a closer look at some of the projects with Total Price = 0. Why is there no cost associated with these projects? Should these projects be removed from the dataset?

#### Project Donations

These fields provide information about the donations made to a particular project.

* total_donations: Total donation amount
* num_donors: Number of unique donors giving to this project
* eligible_double_your_impact_match: True if the project was ever eligible for a Double Your Impact match offer
* eligible_almost_home_match: True if the project was ever eligible for an Almost Home match offer

``` {r}
summary(project$total_donations)
summary(project$num_donors)
draw_histograms(project, log10(project$total_donations), project$funding_status, "Log(Total Donations)", "Funding Status")
```

``` {r}
draw_histograms(project, log10(project$num_donors), project$funding_status, "Log(Number of Donors)", "Funding Status")
```

``` {r}
draw_graphs(project, project$eligible_double_your_impact_match, project$funding_status, "Eligible for Double Your Impact Match", "Funding Status")
```

``` {r}
draw_graphs(project, project$eligible_almost_home_match, project$funding_status, "Eligible for Almost Home Match", "Funding Status")
```

#### Project Status

These fields provide status information and dates for when a project reached their milestones.

* funding_status: Completed, Expired, Live, or Reallocated
* date_posted: Date a project was approved by staff to be listed on the site
* date_completed: Date a project become fully funded
* date_thank_you_packet_mailed: Date that the project went through the final stage of sending out a thank you package to donors
* date_expiration: Date the project was set to expire and be delisted from the site

We have already looked at funding_status. Let's also look at the year the project was posted.

``` {r}
draw_graphs(project, as.factor(format(project$date_posted, '%Y')), project$funding_status, "Year Posted", "Funding Status")
```

#### School Location

The fields describing school location include:

* school_latitude
* school_longitude
* school_city
* school_state
* school_zip: 5 Digit Zip
* school_metro: Urban, Suburban, or Rural
* school_district
* school_county

Let's look at how school metro relates to funding status:
``` {r}
draw_graphs(project, project$school_metro, project$funding_status, "School Metro", "Funding Status")
```


Let's look at percentage of projects completed by state:
``` {r}
project <- project %>% mutate(project_completed = as.integer(funding_status == "completed"))

state_summary <- project %>% group_by(school_state) %>% summarise(value = mean(project_completed) * 100) %>% rename(abb = school_state)

data(state.regions)
state_summary <- left_join(state_summary, state.regions, by="abb")

state_choropleth(state_summary, num_colors = 1, legend = "Percent Completed")
```

#### Additional Properties of Interest

##### New Teacher, New School
The DonorsChoose.org website allows donors to search for projects according to a number of different criteria including "Never before funded teachers" and "Never before funded schools". These properties are not tracked in the original dataset, so let's figure that out now. 

``` {r}
project <- project %>% arrange(date_posted) 
project$teacher_new <- !duplicated(project$X_teacher_acctid)
project$school_new <- !duplicated(project$X_schoolid)

draw_graphs(project, project$teacher_new, project$funding_status, "Never Before Funded Teacher", "Funding Status")
```

``` {r}
draw_graphs(project, project$school_new, project$funding_status, "Never Before Funded School", "Funding Status")

```


