---
title: "Model Building"
output: html_document
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=4)
library(plyr)
library(tidyr)
library(data.table)
library(dplyr)
library(rpart)
library(rpart.plot)
library(caret)
library(e1071)
library(randomForest)
library(tidytext)
library(stringr)
```

## Project Data

#### Importing the file
```{r File Import, cache=TRUE}
project_orig <- fread("opendata_projects000/opendata_projects000.csv", sep = ",", header = FALSE, data.table = FALSE, col.names = c('_projectid', '_teacher_acctid', '_schoolid', 'school_ncesid', 'school_latitude', 'school_longitude', 'school_city', 'school_state', 'school_zip', 'school_metro', 'school_district', 'school_county', 'school_charter', 'school_magnet', 'school_year_round', 'school_nlns', 'school_kipp', 'school_charter_ready_promise', 'teacher_prefix', 'teacher_teach_for_america', 'teacher_ny_teaching_fellow', 'primary_focus_subject', 'primary_focus_area' ,'secondary_focus_subject', 'secondary_focus_area', 'resource_type', 'poverty_level', 'grade_level', 'vendor_shipping_charges', 'sales_tax', 'payment_processing_charges', 'fulfillment_labor_materials', 'total_price_excluding_optional_support', 'total_price_including_optional_support', 'students_reached', 'total_donations', 'num_donors', 'eligible_double_your_impact_match', 'eligible_almost_home_match', 'funding_status', 'date_posted', 'date_completed', 'date_thank_you_packet_mailed', 'date_expiration'))
```

#### Cleaning the data

```{r Clean Data}
project <- tbl_df(project_orig)

project$school_latitude <- as.double(project$school_latitude)
project$school_longitude <- as.double(project$school_longitude)

project$school_city <- as.factor(project$school_city)
project$school_state <- as.factor(project$school_state)
project$school_zip <- as.factor(project$school_zip)
project$school_metro <- as.factor(project$school_metro)
project$school_district <- as.factor(project$school_district)
project$school_county <- as.factor(project$school_county)

project$school_charter <- as.logical(toupper(project$school_charter))
project$school_magnet <- as.logical(toupper(project$school_magnet))
project$school_year_round <- as.logical(toupper(project$school_year_round))
project$school_nlns <- as.logical(toupper(project$school_nlns))
project$school_kipp <- as.logical(toupper(project$school_kipp))
project$school_charter_ready_promise <- as.logical(toupper(project$school_charter_ready_promise))

project$teacher_prefix <- as.factor(project$teacher_prefix)

project$teacher_teach_for_america <- as.logical(toupper(project$teacher_teach_for_america))
project$teacher_ny_teaching_fellow <- as.logical(toupper(project$teacher_ny_teaching_fellow))

project$primary_focus_subject <- as.factor(project$primary_focus_subject)
project$primary_focus_area <- as.factor(project$primary_focus_area)
project$secondary_focus_subject <- as.factor(project$secondary_focus_subject)
project$secondary_focus_area <- as.factor(project$secondary_focus_area)

project$resource_type <- as.factor(project$resource_type)
project$poverty_level <- as.factor(project$poverty_level)
project$grade_level <- as.factor(project$grade_level)

project$vendor_shipping_charges <- as.double(project$vendor_shipping_charges)
project$sales_tax <- as.double(project$sales_tax)
project$payment_processing_charges <- as.double(project$payment_processing_charges)
project$fulfillment_labor_materials <- as.double(project$fulfillment_labor_materials)
project$total_price_excluding_optional_support <- as.double(project$total_price_excluding_optional_support)
project$total_price_including_optional_support <- as.double(project$total_price_including_optional_support)

project$students_reached <- as.integer(project$students_reached)
project$total_donations <- as.integer(project$total_donations)
project$num_donors <- as.integer(project$num_donors)

project$eligible_double_your_impact_match <- as.logical(toupper(as.character(project$eligible_double_your_impact_match)))
project$eligible_almost_home_match <- as.logical(toupper(as.character(project$eligible_almost_home_match)))

project$funding_status <- as.factor(project$funding_status)

project$date_posted <- as.Date(project$date_posted)
project$date_completed <- as.Date(project$date_completed)
project$date_thank_you_packet_mailed <- as.Date(project$date_thank_you_packet_mailed)
project$date_expiration <- as.Date(project$date_expiration)

project$grade_level <- factor(project$grade_level, levels = c("", "Grades PreK-2", "Grades 3-5", "Grades 6-8", "Grades 9-12"))
project$poverty_level <- factor(project$poverty_level, levels = c("", "low poverty", "moderate poverty", "high poverty", "highest poverty"))

```


#### Additional Properties of Interest

#### Focus Area and Subject
The dataset contains values for primary and secondary focus areas and subjects. When
searching through the DonorsChoose.org site, they do not differentiate between primary
and secondary focus areas. Instead, the project is considered to have two focus areas
and two subjects. So we will transform the data to have boolean properties such as
focus_area_applied_learning which will be true if either the primary or secondary
focus area is "Applied Learning."

```{r Focus Area and Subject}
project$focus_area_applied_learning <- as.logical(project$primary_focus_area == "Applied Learning" | project$secondary_focus_area == "Applied Learning")
project$focus_area_health_sports <- as.logical(project$primary_focus_area == "Health & Sports" | project$secondary_focus_area == "Health & Sports")
project$focus_area_history_civics <- as.logical(project$primary_focus_area == "History & Civics" | project$secondary_focus_area == "History & Civics")
project$focus_area_lit_language <- as.logical(project$primary_focus_area == "Literacy & Language" | project$secondary_focus_area == "Literacy & Language")
project$focus_area_math_science <- as.logical(project$primary_focus_area == "Math & Science" | project$secondary_focus_area == "Math & Science")
project$focus_area_music_arts <- as.logical(project$primary_focus_area == "Music & The Arts" | project$secondary_focus_area == "Music & The Arts")
project$focus_area_special_needs <- as.logical(project$primary_focus_area == "Special Needs" | project$secondary_focus_area == "Special Needs")

project$subject_applied_sci <- as.logical(project$primary_focus_subject == "Applied Sciences" | project$secondary_focus_subject == "Applied Sciences")
project$subject_character_ed <- as.logical(project$primary_focus_subject == "Character Education" | project$secondary_focus_subject == "Character Education")
project$subject_civics_gov <- as.logical(project$primary_focus_subject == "Civics & Government" | project$secondary_focus_subject == "Civics & Government")
project$subject_college_career <- as.logical(project$primary_focus_subject == "College & Career Prep" | project$secondary_focus_subject == "College & Career Prep")
project$subject_community <- as.logical(project$primary_focus_subject == "Community Service" | project$secondary_focus_subject == "Community Service")
project$subject_early_dev <- as.logical(project$primary_focus_subject == "Early Development" | project$secondary_focus_subject == "Early Development")
project$subject_economics <- as.logical(project$primary_focus_subject == "Economics" | project$secondary_focus_subject == "Economics")
project$subject_enviro_sci <- as.logical(project$primary_focus_subject == "Environmental Science" | project$secondary_focus_subject == "Environmental Science")
project$subject_esl <- as.logical(project$primary_focus_subject == "ESL" | project$secondary_focus_subject == "ESL")
project$subject_extracurricular <- as.logical(project$primary_focus_subject == "Extracurricular" | project$secondary_focus_subject == "Extracurricular")
project$subject_financial_lit <- as.logical(project$primary_focus_subject == "Financial Literacy" | project$secondary_focus_subject == "Financial Literacy")
project$subject_foreign_lang <- as.logical(project$primary_focus_subject == "Foreign Languages" | project$secondary_focus_subject == "Foreign Languages")
project$subject_gym_fitness <- as.logical(project$primary_focus_subject == "Gym & Fitness" | project$secondary_focus_subject == "Gym & Fitness")
project$subject_health_life_sci <- as.logical(project$primary_focus_subject == "Health & Life Science" | project$secondary_focus_subject == "Health & Life Science")
project$subject_health_wellness <- as.logical(project$primary_focus_subject == "Health & Wellness" | project$secondary_focus_subject == "Health & Wellness")
project$subject_hist_geog <- as.logical(project$primary_focus_subject == "History & Geography" | project$secondary_focus_subject == "History & Geography")
project$subject_literacy <- as.logical(project$primary_focus_subject == "Literacy" | project$secondary_focus_subject == "Literacy")
project$subject_lit_writing <- as.logical(project$primary_focus_subject == "Literature & Writing" | project$secondary_focus_subject == "Literature & Writing")
project$subject_math <- as.logical(project$primary_focus_subject == "Mathematics" | project$secondary_focus_subject == "Mathematics")
project$subject_music <- as.logical(project$primary_focus_subject == "Music" | project$secondary_focus_subject == "Music")
project$subject_nutrition <- as.logical(project$primary_focus_subject == "Nutrition" | project$secondary_focus_subject == "Nutrition")
project$subject_other <- as.logical(project$primary_focus_subject == "Other" | project$secondary_focus_subject == "Other")
project$subject_parent <- as.logical(project$primary_focus_subject == "Parent Involvement" | project$secondary_focus_subject == "Parent Involvement")
project$subject_perform_art <- as.logical(project$primary_focus_subject == "Performing Arts" | project$secondary_focus_subject == "Performing Arts")
project$subject_social_sci <- as.logical(project$primary_focus_subject == "Social Sciences" | project$secondary_focus_subject == "Social Sciences")
project$subject_special_needs <- as.logical(project$primary_focus_subject == "Special Needs" | project$secondary_focus_subject == "Special Needs")
project$subject_team_sports <- as.logical(project$primary_focus_subject == "Team Sports" | project$secondary_focus_subject == "Team Sports")
project$subject_visual_arts <- as.logical(project$primary_focus_subject == "Visual Arts" | project$secondary_focus_subject == "Visual Arts")

```

#### Dealing with Missing Values
In the dataset, we see that the students_reached property includes some zero values and also several NA values. We will take these to be equivalent, and change the NA's to zeroes.

The resource_type property has several missing values, and also a value "Other". Let's consider these as equivalent and change the missing values to "Other."

```{r Missing Values}
project$students_reached[is.na(project$students_reached)] <- as.integer(0)

project$resource_type[project$resource_type == ""] <- "Other"
project$resource_type <- droplevels(project$resource_type)
```

#### Funding status and Date Posted
We will only focus on projects that are either completed (successfully funded) or expired (did not reach funding goal), and remove the live and reallocated projects. To limit the scope of the analysis, we will only focus on projects posted in the last 3 years (2014 onward).

``` {r Funding Status}
project <- project %>% filter(funding_status == "completed" | funding_status == "expired")
project$funding_status <- droplevels(project$funding_status)

project <- subset(project, date_posted >= "2014-01-01")
```


## Essay Data

#### Importing the file
Let's import the file and only keep records that are from 2014 onwards.
```{r Essay Import}
essay_orig <- fread("opendata_essays0002/opendata_essays000.csv", encoding = "UTF-8", sep = ",", header = FALSE, data.table = FALSE, col.names = c('_projectid', '_teacher_acctid', 'title', 'essay'), drop = c(4,5,7,8))

essay_orig <- project %>% subset(select = c("_projectid", "_teacher_acctid")) %>% left_join(essay_orig, by = c("_projectid", "_teacher_acctid"))

```


#### Sentiment Analysis
Since the essay data is too large to analyze all at once, we will process the data in smaller batches. This will include calculating sentiment scores using AFINN and BING Lexicons.

The AFINN lexicon provides a positivity score for each word, from -5 (most negative) to 5 (most positive). We can average the the positivity scores for each essay to get a sentiment score.

The Bing lexicon rates words as positive or negative. We can calculate the difference between the number of positive and negative words in each essay to get a sentiment score.


#### Importing the file
```{r Sentiment}

AFINN <- subset(sentiments, lexicon == "AFINN", select = c(word, score))
bing <- get_sentiments("bing")

sentiment_data <- data.frame()

for (i in 0:44) {
  cat(i)
  firstrow <- 1 + i * 10000
  lastrow <- (i + 1) * 10000
  
  essay_subset <- essay_orig[firstrow:lastrow,] %>% 
    select(`_projectid`, `_teacher_acctid`, essay)

  essay_words <- essay_subset %>%
    unnest_tokens(word, essay)
  
  essay_word_count <- essay_words %>% 
    count(`_projectid`, `_teacher_acctid`) %>% 
    rename(wordcount = n)
    
  essay_words <- essay_words %>%
    filter(!word %in% stop_words$word, str_detect(word, "^[a-z']+$"))
  
  afinn_sentiment <- essay_words %>%
    inner_join(AFINN, by = "word") %>%
    group_by(`_projectid`, `_teacher_acctid`) %>%
    summarize(afinn_score = mean(score))
  
  bing_sentiment <- essay_words %>%
    inner_join(bing, by = "word") %>%
    count(`_projectid`, `_teacher_acctid`, sentiment) %>%
    spread(sentiment, n, fill = 0) %>%
    mutate(bing_count = positive - negative)
  
  essay_sentiment <- essay_word_count %>% 
    left_join(afinn_sentiment, by = c("_projectid", "_teacher_acctid")) %>% 
    left_join(bing_sentiment, by = c("_projectid", "_teacher_acctid")) %>% 
    mutate(bing_score = 100*bing_count/wordcount)

  sentiment_data <- sentiment_data %>% bind_rows(essay_sentiment)
}

project <- project %>% left_join(sentiment_data, by = c("_projectid","_teacher_acctid"))

project$wordcount[is.na(project$wordcount)] <- 0
project$afinn_score[is.na(project$afinn_score)] <- 0
project$bing_score[is.na(project$bing_score)] <- 0

```

```{r  Free up memory}
sort( sapply(ls(),function(x){object.size(get(x))}))
remove(essay_orig)
remove(project_orig)
remove(sentiment_data)
remove(bing_sentiment)
remove(essay_sentiment)
remove(bing)
remove(essay_subset)
remove(AFINN)
remove(essay_word_count)
remove(afinn_sentiment)
remove(essay_words)
gc()
```




## Model Building

The objective of this capstone project is to build a model to predict the chance of getting funding based on certain characteristics of the project and essay. Let's focus on the more recent data from DonorsChoose.org, and use the years 2014-2015 to predict the results from 2016.

```{r Training and Test Sets}
project$completed <- as.logical(project$funding_status == "completed")

train <- subset(project, date_posted >= "2014-01-01" & date_posted <= "2015-12-31")
test <- subset(project, date_posted >= "2016-01-01")
```

For building the model, let's focus on properties of the project that are within the teacher's control. So Focus Subject and Total Price can be set by the teacher, but school location and grade level cannot.

The independent properties will be:
* resource_type
* total_price_excluding_optional_support
* focus_area (7 properties)
* subject (28 properties)
* wordcount
* afinn_score
* bing_score

```{r CART Model}
modelcart <- rpart(funding_status ~ resource_type + total_price_excluding_optional_support + focus_area_applied_learning + focus_area_health_sports + focus_area_history_civics + focus_area_lit_language + focus_area_math_science + focus_area_music_arts + focus_area_special_needs + subject_applied_sci + subject_character_ed + subject_civics_gov + subject_college_career + subject_community + subject_early_dev + subject_economics + subject_enviro_sci + subject_esl + subject_extracurricular + subject_financial_lit + subject_foreign_lang + subject_gym_fitness + subject_health_life_sci + subject_health_wellness + subject_hist_geog + subject_literacy + subject_lit_writing + subject_math + subject_music + subject_nutrition + subject_other + subject_parent + subject_perform_art + subject_social_sci + subject_special_needs + subject_team_sports + subject_visual_arts + wordcount + afinn_score + bing_score, data = train, method = "class", cp = 0.01)

prp(modelcart)

predictiontraincart <- predict(modelcart, type = "class")
table(train$funding_status, predictiontraincart)

predictiontestcart <- predict(modelcart, newdata = test, type = "class")
table(test$funding_status, predictiontestcart)
```

The accuracy of the model on the training set is (232048 + 29037)/360769 = 0.7242

The baseline model (where all projects are predicted as "completed") has an accuracy of (232048 + 23871)/360769 = 0.7094 

The CART model is a modest improvement over the baseline model

The accuracy of the model on the test set is (62132 + 3982)/87117 = 0.7589
The baseline model has an accuracy of (62132 + 4814)/87117 = 0.7685

The CART model does not perform as well as the baseline model on the test set.



```{r Logistic Regression Model}
logmodel <- glm(completed ~ resource_type + total_price_excluding_optional_support + focus_area_applied_learning + focus_area_health_sports + focus_area_history_civics + focus_area_lit_language + focus_area_math_science + focus_area_music_arts + focus_area_special_needs + subject_applied_sci + subject_character_ed + subject_civics_gov + subject_college_career + subject_community + subject_early_dev + subject_economics + subject_enviro_sci + subject_esl + subject_extracurricular + subject_financial_lit + subject_foreign_lang + subject_gym_fitness + subject_health_life_sci + subject_health_wellness + subject_hist_geog + subject_literacy + subject_lit_writing + subject_math + subject_music + subject_nutrition + subject_other + subject_parent + subject_perform_art + subject_social_sci + subject_special_needs + subject_team_sports + subject_visual_arts + wordcount + afinn_score + bing_score, data = train, family = "binomial")

summary(logmodel)
predictiontrainlog <- predict(logmodel, type = "response")
table(train$funding_status, predictiontrainlog < 0.5)

predictiontestlog <- predict(logmodel, newdata = test, type = "response")
table(test$funding_status, predictiontestlog < 0.5)
```

The accuracy of the model on the training set is (247480 + 12551)/360769 = 0.7208

The baseline model (where all projects are predicted as "completed") has an accuracy of 0.7094 

The logistic regression model is a modest improvement over the baseline model

The accuracy of the model on the test set is (65361 + 1762)/87117 = 0.7705
The baseline model has an accuracy of 0.7685

The logistic regression model performs slightly better on the test set than the baseline model.
